\section{\label{background}Background}

\subsection{\label{sfm}The Social Force Model}

Our research was carried out in the context of the ``social force model'' (SFM) 
proposed by Helbing and co-workers \cite{Helbing1}. This model states that human 
motion is caused by the desire of people to reach a certain destination, as 
well as other environmental factors. The pedestrians behavioral pattern in a 
crowded environment can be modeled by three kind of forces: the ``desire 
force'', the ``social force'' and the ``granular force''.\\

The ``desire force'' represents the pedestrian's own desire to reach a 
specific target position at a desired velocity $v_d$. But, in order to reach 
the desired target, he (she) needs to accelerate (decelerate) from his (her) 
current velocity $\mathbf{v}^{(i)}(t)$. This acceleration (or deceleration) 
represents a ``desire force'' since it is motivated by his (her) own 
willingness. The corresponding expression for this forces is 

\begin{equation}
        \mathbf{f}_d^ {(i)}(t) =  
m_i\,\displaystyle\frac{v_d^{(i)}\,\mathbf{e}_d^
{(i)}(t)-\mathbf{v}^{(i)}(t)}{\tau} \label{desired}
\end{equation}

where $m_i$ is the mass of the pedestrian $i$. $\mathbf{e}_d$ 
corresponds to the unit vector pointing to the target position and $\tau$ is a 
constant related to the relaxation time needed to reach his (her) desired 
velocity. For simplicity, we assume that 
$v_d$ remains constant during the entire process and is the same for all 
individuals, but $\mathbf{e}_d$ changes according to the current position of the 
pedestrian. Detailed values for $m_i$ and $\tau$ can be found in 
Refs.~\cite{Helbing1,Dorso3}. \\

The ``social force'' represents the psychological tendency of two pedestrians,  
say $i$ and $j$, to stay away from each other by a repulsive interaction force 

\begin{equation}
        \mathbf{f}_s^{(ij)} = A_i\,e^{(R_{ij}-r_{ij})/B_i}\mathbf{n}_{ij} 
        \label{social}
\end{equation}

where $(ij)$ means any pedestrian-pedestrian pair, or pedestrian-wall 
pair. $A_i$ and $B_i$ are fixed values, $r_{ij}$ is the distance between  the 
center of mass of the pedestrians $i$ and $j$ and the distance $R_{ij}=R_i+R_j$ 
is the sum of the pedestrians radius. $\mathbf{n}_{ij}$ means the unit vector in 
the $\vec{ji}$ direction.\\

Any two pedestrians touch each other if their distance $r_{ij}$ is smaller than 
$R_{ij}$. Analogously, a pedestrian touch a wall if his/her distance $r_{ij}$ to the wall is smaller than $R_i$. In these cases, an additional force is included in the model, called the ``granular force''(\textit{i.e.} friction force). This force is considered be a linear function of the relative (tangential) velocities of the contacting individuals. In the case of the friction exerted by the wall, the force is a linear function of the tangential velocity. % Mejorar esta oracion!
Its mathematical expression reads 

\begin{equation}
        \mathbf{f}_g^{(ij)} = 
\kappa\,(R_{ij}-r_{ij})\,\Theta(R_{ij}-r_{ij})\,\Delta
\mathbf{v}^{(ij)}\cdot\mathbf{t}_{ij} 
        \label{granular}
\end{equation}

where $\kappa$ is the friction coefficient. The function 
$\Theta(R_{ij}-r_{ij})$ is zero when its argument is negative (that is, 
$R_{ij}<r_{ij}$) and equals unity for any other case (Heaviside function). 
$\Delta\mathbf{v}^{(ij)}\cdot\mathbf{t}_{ij}$ represents the difference between 
the tangential velocities of the sliding bodies (or between the individual and 
the walls).   

The above forces actuate on the pedestrians dynamics by changing his (her) 
current velocity. The equation of motion for pedestrian $i$ reads

\begin{equation}
m_i\,\displaystyle\frac{d\mathbf{v}^{(i)}}{dt}=\mathbf{f}_d^{(i)}
+\displaystyle\sum_{j=1}^{N}\displaystyle\mathbf{f}_s^{(ij)}
+\displaystyle\sum_ {
j=1}^{N}\mathbf{f}_g^{(ij)}\label{eq_mov}
\end{equation}

where the subscript $j$ represents all the other pedestrians 
(excluding $i$) and the walls.\\

In the original model, there is no distinction between the friction coefficient of pedestrian-pedestrian interaction and pedestrian-wall interaction. Both interactions are modeled with the constant parameter $\kappa$. In this paper we analyze situations in which the friction coefficient may take different values. We define $\kappa_i$ and $\kappa_w$ as the friction coefficient related to the pedestrian-pedestrian interaction and the pedestrian-wall interaction respectively. 

\subsection{\label{fundamental-diagram} Fundamental Diagram}

Inspired in the vehicular traffic dynamic studies, many researches of pedestrian dynamics focus its attention on the relation between the flow and density of a moving crowd. This analysis is called fundamental diagram and it has become one of the most common ways to characterize the pedestrians dynamics in unidirectional and bidirectional flows \cite{fruin1,mori1,polus1,seyfried1,jelic1}. \\

In this paper we use the mathematical expressions defined by Helbing et al. in Ref.~\cite{helbing3} as regards to the fundamental diagram analysis. Thus, the local density at place $\vec{r}=(x,y)$ and time $t$ is defined as

\begin{equation}
\rho(\vec{r},t)=\sum_{j}f(\vec{r}_j(t)-\vec{r}) \label{ec-density}
\end{equation}

where function $f(\vec{r}_j(t)-\vec{r})$ is a Gaussian distance-dependent weight function defined as

\begin{equation}
f(\vec{r}_j-\vec{r})=\frac{1}{\pi R^2}\exp[-\left \| \vec{r}_j-\vec{r} \right \|^2/R^2] \label{ec-f}
\end{equation}

$\vec{r}_j(t)$ are the positions of the pedestrians $j$ in the surrounding of $\vec{r}$ and R is a measurement parameter. 
The local speeds have been defined via the weighted average  

\begin{equation}
\vec{V}(\vec{r},t)=\frac{\sum_j \vec{v}_jf(\vec{r}_j(t)-\vec{r}) }{\sum_j f(\vec{r}_j(t)-\vec{r}) } \label{ec-v}
\end{equation}

while flow have been determined according to the fluid-dynamic formula

\begin{equation}
\vec{J}(\vec{r},t)=\rho(\vec{r},t)\vec{V}(\vec{r},t) \label{ec-flow}
\end{equation}


\subsection{\label{granular-cluster} Clustering structures}

Groups of pedestrians can be defined as the set of individuals that for any member of the group (say, $i$) there exists at least another member belonging to the same group ($j$) in contact with the former. Thus, we define a ``granular cluster'' ($C_g$) following the mathematical formula given in Ref.~\cite{Dorso1}

\begin{equation}
C_g:P_i~\epsilon~ C_g \Leftrightarrow \exists~ j~\epsilon~C_g / r_{ij} < (R_i+R_j) \label{ec-cluster}
\end{equation}


where ($P_i$) indicate the ith pedestrian and $R_i$ is his/her radius (shoulders width). That means, $C_g$ is a
set of pedestrians that interact not only with social and desired force, but also with granular forces (\textit{i.e.} friction forces).
The size of the cluster is defined as the number of pedestrians that compose it. The fraction of clusterized individuals is defined as the quotient between clustered individuals with respect to the total number of individuals in the crowd. In Section \ref{clusters} we analyze the clustering structures in terms of these two observables. 